---
title: "Untitled"
author: "Erika Mudrak"
date: "Friday, September 05, 2014"
output: html_document
---

```{r load libraries, echo=FALSE, results='hide', message=FALSE}


library(nlme)
#library(spatstat)
library(devtools)  #do once
install_github("elmR", "emudrak")  #do once, equiv to install.packages()
library(elmR)
#this is a package I wrote with following transformation functions----
logit= function (x) 
{
    if (sum(x[x != 0], na.rm = TRUE) == 0) 
        eps = 0
    else eps = min(x[x != 0], na.rm = TRUE)
    x[x == 0] = 0.5 * eps
    x[x == 1] = 1 - 0.5 * eps
    return(log(x/(1 - x)))
}
inv_logit=function (alpha) 
{
    alpha[alpha > 709] = 709
    return(exp(alpha)/(exp(alpha) + 1))
}
logw0=function (x) 
{
    c = trunc(log(min(x[x > 0], na.rm = T)))
    d = exp(c)
    return(log(x + d) - c)
}
# end (mini) package  -----

#setwd("C:/Users/elm26/Documents/SERDPproject/VegSpatialModeling/VegSpatialModeling")
```



```{r Read in and munge data , echo=FALSE, results='hide'}
# PICK A DESERT
#DESERT="Sonoran"
DESERT="Mojave"
YEAR="2013Xtra"

if((DESERT=="Mojave") & (YEAR=="2013Xtra")){
  CensusData=read.csv("../Mojave2013Intense24Shrubs.csv", blank.lines.skip=TRUE)
  CensusData$Invasives=CensusData$Erodium+CensusData$Schismus+CensusData$Bromus
  Shrubs=read.csv("../CALarreaVolume.csv")
  Shrubs=subset(Shrubs, ShrubID %in% as.character(1:168))
  Shrubs$Shrub=as.numeric(as.character(Shrubs$ShrubID)) 
}

Census=merge(Shrubs[,c("Shrub","Area_Bot","Vol_Bot","Vol_Stem")], CensusData,  by.y="Shrub", by.x="Shrub")

Census$Area_Bot=Census$Area_Bot/10000
Census$Vol_Bot=Census$Vol_Bot/1000000
Census$Vol_Stem=Census$Vol_Stem/1000000
Census$DistClass=cut(Census$MHcode, breaks=c(0,2,3,4), labels=c("UC","ON","OF"))


Census=Census[order(Census[,"Shrub"], Census[,"MHcode"]), ]
names(Census)[1]="Shrub"


 
```




```{r Choose Target, results='hide', echo=FALSE}
# Chose target species to model this time --------------
TargSpec="NatCover"; ylabel="% Cover" 
#TargSpec="Erodium"; ylabel="# Plants" 

Census$Target=round(Census[,TargSpec]) #Round to deal with issue of 0.1 percent cover values
Census$LogitTarg=logit(Census$Target/100)
Census$TargPres=ceiling(Census$Target/100) #Make a new binary variable showing zero or not

write.csv(mean(Census$Target[Census$DistClass=="OF"])/100, file="Interspace.csv", row.names=F)
```

Using only data with Ambient Rainfall levels, split into testing and training data, stratified by Fire and TranDir treatments. 

```{r Training/Testing Data, results='hide', echo=FALSE}
# Make training and Testing Data- ------------
# balance out Fire, Rain and TranDir, 7 of each shrub 
#  Withhold two in each category for testing
set.seed(81)
BurnAmbN=sample(unique(subset(Census, (Fire=="B")&(Rain=="A")&(TranDir=="N"))$Shrub), 2)
BurnAmbS=sample(unique(subset(Census, (Fire=="B")&(Rain=="A")&(TranDir=="S"))$Shrub), 2)
BurnDrtN=sample(unique(subset(Census, (Fire=="B")&(Rain=="D")&(TranDir=="N"))$Shrub), 2)
UnBurnAmbN=sample(unique(subset(Census, (Fire=="UB")&(Rain=="A")&(TranDir=="N"))$Shrub), 2)
UnBurnAmbS=sample(unique(subset(Census, (Fire=="UB")&(Rain=="A")&(TranDir=="S"))$Shrub), 2)
UnBurnDrtN=sample(unique(subset(Census, (Fire=="UB")&(Rain=="D")&(TranDir=="N"))$Shrub), 2)
TestShrubs=c(BurnAmbN, BurnAmbS, BurnDrtN, UnBurnAmbN,  UnBurnAmbS,  UnBurnDrtN)

TestShrubs
#58 131  99  14 124 120 122  80  93  63  69  26

#Use only Ambient Shrubs

Census.Test=Census[Census$Shrub %in% TestShrubs & Census$Rain=="A",]
Census.Train=Census[!(Census$Shrub %in% TestShrubs) & Census$Rain=="A",]


```


This time using `r TargSpec`

Response seems to be zero-inflated

```{r ExplorDatAnalysis, echo=FALSE, fig.height=3}
par(mfrow=c(1,2))
with(Census.Train, hist(logw0(Target/100), breaks= 40, xlab=paste("log ", ylabel), main=NA))

with(Census.Train,hist(logit(Target/100), breaks= 40, xlab=paste("logit ",ylabel), main=NA))

```



```{r Plot With transformed response, echo=FALSE, message=FALSE, warning=FALSE, fig.height=3.5}

# Setup for elaborate plotting 
colorset=sample(rainbow(168))
firecols=c("gray40", "green")
dircols=c("red","blue")
dirlty=c(1,2)
firepch=c(16,1)
areawts=c(1,2,3)



blah=boxplot(LogitTarg~DistClass*Fire*TranDir, data=Census.Train, axes=FALSE, frame=T, col=rep(firecols,each=3, times=2), ylab="Percent Cover")
Axis(at=1:12, labels=substr(blah$names,1,2), side=1)
Axis(side=2, at=logit(seq(0,0.3, by=0.01)), labels=seq(0,0.3, by=0.01),cex.axis=0.8)
mtext(rep(levels(Census.Train$Fire),2), line=2, side=1, at=c(2,5,8,11))
axis( at=c(3.5, 6.5, 9.5), side=1, label=F, tck=-0.2 )
mtext(levels(Census.Train$TranDir), line=3, side=1, at=c(3.5, 9.5))
axis( at=c( 6.5), side=1, label=F, tck=-0.3 )

library(plyr)
Agr.Train=ddply(Census.Train, .(Shrub,DistClass, Fire,Area_Bot), function(x) c(meanCover=mean(x$Target), stderr=sqrt(var(x$Target)/length(x$Target))))

boxplot(meanCover~ DistClass, data=Agr.Train, col="white", lty="blank", ylim=c(0,32), outline=F, ylab="Percent Cover")
#Add error bars, but messy
d_ply(Agr.Train, .(Shrub), function(x) {
    lines(x$DistClass, x$meanCover, col=firecols[x$Fire]) 
    xes=jitter(1:3, amount=0.02)
    arrows(xes, x$meanCover,xes, x$meanCover+x$stderr, angle=90, length=0.02, col=firecols[x$Fire])
    arrows(xes, x$meanCover,xes, x$meanCover-x$stderr, angle=90, length=0.02, col=firecols[x$Fire])
    })

#Set scale ranges for this target
xplotmax=max(Census.Train[,"PlotDist"], na.rm=TRUE)
xplotmax=3
#xplotmax=max(Census.Train[,"MHcode"], na.rm=TRUE)
yplotmax=max(Census.Train[,"Target"])

# 
# plotscheme="fire"
# # plotscheme="direction"
# 
# if (plotscheme=="fire"){
#     thesecols=firecols; names=levels(Census.Train$Fire)
#     }else 
# if (plotscheme=="rain"){
#       thesecols=watercols ; names=levels(Census.Train$Rain)
#     }else 
# if (plotscheme=="direction"){
#       thesecols=dircols; names=levels(Census.Train$TranDir)  
#     }
# legend(2.1,0.90* yplotmax, levels(Census.Train$Fire), col=thesecols, lwd=2, lty=1, bty="n") 

```


Because of the zero-inflation, fit a two-stage model.  

### Hurdle Model------------ 
First fit "hurdle" model, i.e. a presence/absense model indicating whether the plot has plants or not. 
Using only plots from Canopy Dripline and outward (i.e. MHCode > 1)
Model fitting proccess hidden, only final model results shown.

```{r Logistic modeling, echo=FALSE, message=FALSE, warning=FALSE}
library(lme4)
library(lmerTest)


hurdle1.lme=glmer(TargPres~(DistClass+Area_Bot+Fire+TranDir)^2 + (1|Shrub), family=binomial, data=Census.Train)
summary(hurdle1.lme)
drop1(hurdle1.lme, test="Chisq")

hurdle2.lme=glmer(TargPres~DistClass+Area_Bot+Fire+TranDir + DistClass:Fire +DistClass:TranDir+Area_Bot:Fire + Area_Bot:TranDir + Fire:TranDir  + (1|Shrub), family=binomial, data=Census.Train)
summary(hurdle2.lme)
drop1(hurdle2.lme, test="Chisq")

hurdle3.lme=glmer(TargPres~DistClass+Area_Bot+Fire+TranDir +DistClass:TranDir+Area_Bot:Fire + Area_Bot:TranDir + Fire:TranDir  + (1|Shrub), family=binomial, data=Census.Train)
drop1(hurdle3.lme, test="Chisq")

hurdle4.lme=glmer(TargPres~DistClass+Area_Bot+Fire+TranDir +DistClass:TranDir+Area_Bot:Fire + (1|Shrub), family=binomial, data=Census.Train)
drop1(hurdle4.lme, test="Chisq")

hurdle5.lme=glmer(TargPres~DistClass+Area_Bot+Fire+TranDir +DistClass:TranDir + (1|Shrub), family=binomial, data=Census.Train)
summary(hurdle5.lme)
drop1(hurdle5.lme, test="Chisq")

hurdle6.lme=glmer(TargPres~DistClass+Fire+TranDir +DistClass:TranDir + (1|Shrub), family=binomial, data=Census.Train)
 drop1(hurdle6.lme, test="Chisq")


```

What are the fixed effects?  
```{r look at fixed effects}
# Fixed effects
fixef(hurdle6.lme)

```

Apply model to testing data to see how we did.
```{r Try with Testing Data, echo=FALSE}

#Not stochastic, uses population level data for previously unobserved levels
Census.Test$HurdPred=NA
Census.Test$HurdPred=predict(hurdle6.lme,  newdata=Census.Test, allow.new.levels=TRUE )

plot(jitter(TargPres,factor=0.2)~inv_logit(HurdPred), data=Census.Test, xlab="Probability of presence", ylab="Observed presence (jittered)" ) 

# 
# #Try to predict by equation....
# #logitP
# Eq=with(Census.Test,   3.748875457      -0.000842722*PlotDist      -5.686315700*(Rain=="D")       -4.259545202  *(TranDir=="S") +     0.025186048 *PlotDist*(TranDir=="S"))
# Fn=predict(hurdle12.lme,  newdata=Census.Test, allow.new.levels=TRUE )
# # plot(Eq, Fn, abline(c(0,1)))


```
Not bad!  

### Now model presence-only ------------
Now fit a model where if present, how much cover a is in a plot?  
Using only non-empty plots from Canopy Dripline and outward (i.e. MHCode > 1 & Census.Test$Target>0)
Model fitting proccess hidden, only final model results shown.

```{r presence-only lme, echo=FALSE}
library(lme4)
library(lmerTest)
# 
#


# linmod1.lme=lmer(LogitTarg~(DistClass+Area_Bot+Fire+TranDir)^2 + (1|Shrub), data=Census.Train[Census.Train$Target>0,])
# summary(linmod1.lme)
# drop1(linmod1.lme, test="Chisq")
# 
# linmod2.lme=lmer(LogitTarg~DistClass+Area_Bot+Fire+TranDir + DistClass:Fire +DistClass:TranDir+Area_Bot:Fire + Fire:TranDir  + (1|Shrub), data=Census.Train[Census.Train$Target>0,])
# summary(linmod2.lme)
# drop1(linmod2.lme, test="Chisq")
# 
# linmod3.lme=lmer(LogitTarg~DistClass+Area_Bot+Fire+TranDir + DistClass:TranDir+Area_Bot:Fire + Fire:TranDir  + (1|Shrub), data=Census.Train[Census.Train$Target>0,])
# summary(linmod3.lme)
# drop1(linmod3.lme, test="Chisq")

linmod4.lme=lmer(LogitTarg~DistClass+Area_Bot+Fire+TranDir + DistClass:TranDir+Area_Bot:Fire + (1|Shrub), data=Census.Train[Census.Train$Target>0,])
summary(linmod4.lme)
# drop1(linmod4.lme, test="Chisq")



anova(linmod4.lme)


```
Model Diagnostics

Residual plot looks ok, residuals a little leptokurtic, but at least centered
```{r Model diagnostics, echo=FALSE, fig.height=3.5}
par(mfrow=c(1,3))
plot(fitted(linmod4.lme), resid(linmod4.lme), abline(c(0,0), lty=2))
qqnorm(resid(linmod4.lme))
abline(c(0,1))
hist(resid(linmod4.lme), breaks=40, freq=F)
curve(dnorm(x, mean=mean(resid(linmod4.lme)), sd=sd(resid(linmod4.lme))), add=TRUE)
#Go with this one for now???


```

```{r Presence Only fixed effects}
# Fixed effects
fixef(linmod4.lme)

```
Try on Testing Data. 
Not stochastic, uses population level data for previously unobserved levels

```{r Testing Data, echo=FALSE}


Census.Test$PresOnlyPred=NA
Census.Test[Census.Test$MHcode>1 & Census.Test$Target>0,"PresOnlyPred"]=predict(linmod4.lme,  newdata=Census.Test[Census.Test$MHcode>1& Census.Test$Target>0,], allow.new.levels=TRUE )

plot(Target/100~inv_logit(PresOnlyPred), data=Census.Test[Census.Test$MHcode>1& Census.Test$Target>0,], xlab="Predicted Percent Cover", ylab="Observed percent Cover" ) 
abline(0,1, lty=2)



```

Not Great.

```{r Save models, echo=FALSE, message=FALSE, results='hide', warning=FALSE}
GoodPresenceModel=as.data.frame(t(fixef(hurdle6.lme)))
names(GoodPresenceModel)[1]="Int"
names(GoodPresenceModel)[which(names(GoodPresenceModel)=="PlotDist:TranDirS")]="Dist.DirS"
write.csv(GoodPresenceModel, "ErodPresenceModel.csv", row.names=F)
GoodCoverModel=as.data.frame(t(fixef(linmod4.lme)))
names(GoodCoverModel)[1]="Int"
names(GoodCoverModel)[which(names(GoodCoverModel)=="PlotDist:Area_Bot")]="Dist.Area"
names(GoodCoverModel)[which(names(GoodCoverModel)=="Area_Bot:FireUB")]="Area.FireUB"
write.csv(GoodCoverModel, "ErodCoverModel.csv", row.names=F)
        

```

What are the predicted probabilities and covers for a plot of a given fire status or direction from shrub?

```{r plot predict results, echo=FALSE, warning=FALSE}

NewShrubs=expand.grid( DistClass=c("UC","ON","OF"), Fire=c("B","UB"), TranDir=c("N","S"), Area_Bot=mean(Census$Area_Bot), Shrub=1)

NewShrubs$PredHurd=predict(hurdle6.lme,  newdata=NewShrubs, allow.new.levels=TRUE )

NewShrubs$PredCov=predict(linmod4.lme,  newdata=NewShrubs, allow.new.levels=TRUE )


boxplot(inv_logit(PredHurd)~ DistClass, data=NewShrubs, col="white", lty="blank", outline=F, ylab="predicted probability of plant presence")
d_ply(NewShrubs, .(Fire,TranDir, Area_Bot), function(x) {
    lines(x$DistClass, inv_logit(x$PredHurd), col=firecols[x$Fire], lty=dirlty[x$TranDir]) 
    })
legend(0.5,1.2, levels(NewShrubs$Fire), col=firecols, lty=1, bty="n", horiz=T, xpd=T)
legend(2.5,1.2, levels(NewShrubs$TranDir), lty=dirlty, bty="n", horiz=T, xpd=T)


boxplot(inv_logit(PredCov)~ DistClass, data=NewShrubs, col="white", lty="blank", outline=F, ylab="predicted cover")
#Add error bars, but messy
d_ply(NewShrubs, .(Fire,TranDir), function(x) {
    lines(x$DistClass, inv_logit(x$PredCov), col=firecols[x$Fire], lty=dirlty[x$TranDir]) 
    })

legend(0.5,0.14, levels(NewShrubs$Fire), col=firecols, lty=1, bty="n", horiz=T, xpd=T)
legend(2.5,0.14, levels(NewShrubs$TranDir), lty=dirlty, bty="n", horiz=T, xpd=T)



```

