---
title: "Untitled"
author: "Erika Mudrak"
date: "Friday, September 05, 2014"
output: html_document
---

```{r load libraries, echo=FALSE, results='hide', message=FALSE}


library(nlme)
#library(spatstat)
library(devtools)  #do once
install_github("elmR", "emudrak")  #do once, equiv to install.packages()
library(elmR)
#this is a package I wrote with following transformation functions----
logit= function (x) 
{
    if (sum(x[x != 0], na.rm = TRUE) == 0) 
        eps = 0
    else eps = min(x[x != 0], na.rm = TRUE)
    x[x == 0] = 0.5 * eps
    x[x == 1] = 1 - 0.5 * eps
    return(log(x/(1 - x)))
}
inv_logit=function (alpha) 
{
    alpha[alpha > 709] = 709
    return(exp(alpha)/(exp(alpha) + 1))
}
logw0=function (x) 
{
    c = trunc(log(min(x[x > 0], na.rm = T)))
    d = exp(c)
    return(log(x + d) - c)
}
# end (mini) package  -----

#setwd("C:/Users/elm26/Documents/SERDPproject/VegSpatialModeling/VegSpatialModeling")
```



```{r Read in and munge data , echo=FALSE, results='hide'}
# PICK A DESERT
#DESERT="Sonoran"
DESERT="Mojave"
YEAR="2013Xtra"

if((DESERT=="Mojave") & (YEAR=="2013Xtra")){
  CensusData=read.csv("../Mojave2013Intense24Shrubs.csv", blank.lines.skip=TRUE)
  CensusData$Invasives=CensusData$Erodium+CensusData$Schismus+CensusData$Bromus
  Shrubs=read.csv("../CALarreaVolume.csv")
  Shrubs=subset(Shrubs, ShrubID %in% as.character(1:168))
  Shrubs$Shrub=as.numeric(as.character(Shrubs$ShrubID)) 
}

Census=merge(Shrubs[,c("Shrub","Area_Bot","Vol_Bot","Vol_Stem")], CensusData,  by.y="Shrub", by.x="Shrub")

Census$Area_Bot=Census$Area_Bot/10000
Census$Vol_Bot=Census$Vol_Bot/1000000
Census$Vol_Stem=Census$Vol_Stem/1000000
Census$DistClass=cut(Census$MHcode, breaks=c(0,2,3,4), labels=c("UC","ON","OF"))


Census=Census[order(Census[,"Shrub"], Census[,"MHcode"]), ]
names(Census)[1]="Shrub"


 
```




```{r Choose Target, results='hide', echo=FALSE}
# Chose target species to model this time --------------
TargSpec="ErodCover"; ylabel="% Cover" 
#TargSpec="Erodium"; ylabel="# Plants" 
TargSpec="SchisCover"; ylabel="% Cover" 

Census$Target=round(Census[,TargSpec]) #Round to deal with issue of 0.1 percent cover values
Census$LogitTarg=logit(Census$Target/100)
Census$TargPres=ceiling(Census$Target/100) #Make a new binary variable showing zero or not

write.csv(mean(Census$Target[Census$DistClass=="OF"])/100, file="SchisInterspace.csv", row.names=F)
```



```{r Training/Testing Data, results='hide', echo=FALSE}
# Make training and Testing Data- ------------
# balance out Fire, Rain and TranDir, 7 of each shrub 
#  Withhold two in each category for testing
set.seed(81)
BurnAmbN=sample(unique(subset(Census, (Fire=="B")&(Rain=="A")&(TranDir=="N"))$Shrub), 2)
BurnAmbS=sample(unique(subset(Census, (Fire=="B")&(Rain=="A")&(TranDir=="S"))$Shrub), 2)
BurnDrtN=sample(unique(subset(Census, (Fire=="B")&(Rain=="D")&(TranDir=="N"))$Shrub), 2)
UnBurnAmbN=sample(unique(subset(Census, (Fire=="UB")&(Rain=="A")&(TranDir=="N"))$Shrub), 2)
UnBurnAmbS=sample(unique(subset(Census, (Fire=="UB")&(Rain=="A")&(TranDir=="S"))$Shrub), 2)
UnBurnDrtN=sample(unique(subset(Census, (Fire=="UB")&(Rain=="D")&(TranDir=="N"))$Shrub), 2)
TestShrubs=c(BurnAmbN, BurnAmbS, BurnDrtN, UnBurnAmbN,  UnBurnAmbS,  UnBurnDrtN)

TestShrubs
#58 131  99  14 124 120 122  80  93  63  69  26

#Use only Ambient Shrubs

Census.Test=Census[Census$Shrub %in% TestShrubs & Census$Rain=="A",]
Census.Train=Census[!(Census$Shrub %in% TestShrubs) & Census$Rain=="A",]


```


This time using `r TargSpec`

Response seems to be zero-inflated

```{r ExplorDatAnalysis, echo=FALSE, fig.height=3}
par(mfrow=c(1,2))
with(Census.Train, hist(logw0(Target/100), breaks= 40, xlab=paste("log ", ylabel), main=NA))

with(Census.Train,hist(logit(Target/100), breaks= 40, xlab=paste("logit ",ylabel), main=NA))

mean(CensusData$ErodCover[CensusData$MHcode>=3])


```



```{r Plot With transformed response, echo=FALSE}

# Setup for elaborate plotting 
colorset=sample(rainbow(168))
firecols=c("gray40", "green")
dircols=c("red","blue")
firepch=c(16,1)



boxplot(LogitTarg~DistClass*Fire*TranDir, data=Census.Train)


#Set scale ranges for this target
xplotmax=max(Census.Train[,"PlotDist"], na.rm=TRUE)
#xplotmax=max(Census.Train[,"MHcode"], na.rm=TRUE)
yplotmax=max(Census.Train[,"LogitTarg"])


plotscheme="fire"
# plotscheme="direction"

with(Census.Train, plot(c(1,xplotmax), range(Census.Train$LogitTarg) , pch=NA, main=paste(DESERT, YEAR, TargSpec), ylab=paste("Logit transformmed ", ylabel), xlab="Distance from Shrub Stem" ))
for (i in unique(Census.Train$Shrub)) {  # i=ShrubNumber
    theseplots=Census.Train[Census.Train$Shrub==i,]
    if (plotscheme=="fire") mycol=firecols[theseplots$Fire[1] ] else    
    if (plotscheme=="direction") mycol=dircols[theseplots$TranDir[1] ]
   lines(theseplots$PlotDist, theseplots$LogitTarg,   
#    lines(theseplots$MHcode, theseplots$LogitTarg,        
        pch=20,lwd=2, type="o", lty=1,
        col=mycol  )
} # end of i loop
if (plotscheme=="fire"){
    thesecols=firecols; names=levels(Census.Train$Fire)
    }else 
if (plotscheme=="rain"){
      thesecols=watercols ; names=levels(Census.Train$Rain)
    }else 
if (plotscheme=="direction"){
      thesecols=dircols; names=levels(Census.Train$TranDir)  
    }
legend(200,0.90* yplotmax, levels(Census.Train$Fire), col=thesecols, lwd=2, lty=1, bty="n") 

```


### Hurdle Model------------ 
First fit "hurdle" model, i.e. a presence/absense model indicating whether the plot has plants or not. 
Using only plots from Canopy Dripline and outward (i.e. MHCode > 1)
Model fitting proccess hidden, only final model results shown.

```{r Logistic modeling, echo=FALSE, message=FALSE}
library(lme4)
library(lmerTest)


hurdle1.lme=glmer(TargPres~(DistClass+Area_Bot+Fire+TranDir)^2 + (1|Shrub), family=binomial, data=Census.Train,control = glmerControl(optimizer = "bobyqa"))
summary(hurdle1.lme)
drop1(hurdle1.lme, test="Chisq")

hurdle2.lme=glmer(TargPres~(DistClass+Area_Bot+Fire+TranDir)^2 - DistClass:TranDir + (1|Shrub), family=binomial, data=Census.Train,control = glmerControl(optimizer = "bobyqa"))
summary(hurdle2.lme)
drop1(hurdle2.lme, test="Chisq")


hurdle3.lme=glmer(TargPres~(DistClass+Area_Bot+Fire+TranDir)^2 - DistClass:TranDir - DistClass:Area_Bot + (1|Shrub), family=binomial, data=Census.Train,control = glmerControl(optimizer = "bobyqa"))
summary(hurdle3.lme)
drop1(hurdle3.lme, test="Chisq")

hurdle4.lme=glmer(TargPres~(DistClass+Area_Bot+Fire+TranDir)^2 - DistClass:TranDir - DistClass:Area_Bot - DistClass:Fire + (1|Shrub), family=binomial, data=Census.Train,control = glmerControl(optimizer = "bobyqa"))
summary(hurdle4.lme)
drop1(hurdle4.lme, test="Chisq")

hurdle5.lme=glmer(TargPres~ Area_Bot + Fire + TranDir + Area_Bot:Fire      + Area_Bot:TranDir + Fire:TranDir+  (1|Shrub), family=binomial, data=Census.Train,control = glmerControl(optimizer = "bobyqa"))
summary(hurdle5.lme)
drop1(hurdle5.lme, test="Chisq")

hurdle6.lme=glmer(TargPres~ Area_Bot + Fire + TranDir + Area_Bot:Fire      + Area_Bot:TranDir + Fire:TranDir+  (1|Shrub), family=binomial, data=Census.Train,control = glmerControl(optimizer = "bobyqa"))
summary(hurdle6.lme)
drop1(hurdle6.lme, test="Chisq")


```

What are the fixed effects?  
```{r look at fixed effects}
# Fixed effects
fixef(hurdle6.lme)

```


```{r Try with Testing Data, echo=FALSE}

#Not stochastic, uses population level data for previously unobserved levels
Census.Test$HurdPred=NA
Census.Test$HurdPred=predict(hurdle6.lme,  newdata=Census.Test, allow.new.levels=TRUE )

plot(jitter(TargPres,factor=0.2)~inv_logit(HurdPred), data=Census.Test, xlab="Probability of presence", ylab="Observed presence (jittered)" ) 


#Try to predict by equation....
#logitP
Eq=with(Census.Test,   3.748875457      -0.000842722*PlotDist      -5.686315700*(Rain=="D")       -4.259545202  *(TranDir=="S") +     0.025186048 *PlotDist*(TranDir=="S"))
Fn=predict(hurdle12.lme,  newdata=Census.Test, allow.new.levels=TRUE )
# plot(Eq, Fn, abline(c(0,1)))


```
Not bad!  

### Now model presence-only ------------
Now fit a model where if present, how much cover a is in a plot?  
Using only non-empty plots from Canopy Dripline and outward (i.e. MHCode > 1 & Census.Test$Target>0)
Model fitting proccess hidden, only final model results shown.

```{r presence-only lme, echo=FALSE}
library(lme4)
library(lmerTest)
# 
#


linmod1.lme=lmer(LogitTarg~(DistClass+Area_Bot+Fire+TranDir)^3 + (1|Shrub), data=Census.Train[Census.Train$Target>0,])
summary(linmod1.lme)
drop1(linmod1.lme, test="Chisq")

linmod2.lme=lmer(LogitTarg~(DistClass+Area_Bot+Fire+TranDir)^3 - Area_Bot:Fire:TranDir  + (1|Shrub), data=Census.Train[Census.Train$Target>0,])
summary(linmod2.lme)
drop1(linmod2.lme, test="Chisq")


linmod3.lme=lmer(LogitTarg~(DistClass+Area_Bot+Fire+TranDir)^3 - Area_Bot:Fire:TranDir - DistClass:Fire:TranDir + (1|Shrub), data=Census.Train[Census.Train$Target>0,])
summary(linmod3.lme)
drop1(linmod3.lme, test="Chisq")

linmod4.lme=lmer(LogitTarg~(DistClass+Area_Bot+Fire+TranDir)^3 - Area_Bot:Fire:TranDir - DistClass:Fire:TranDir - Fire:TranDir+ (1|Shrub), data=Census.Train[Census.Train$Target>0,])
summary(linmod4.lme)
drop1(linmod4.lme, test="Chisq")

linmod5.lme=lmer(LogitTarg~(DistClass+Area_Bot+Fire+TranDir)^3 - Area_Bot:Fire:TranDir - DistClass:Fire:TranDir - Fire:TranDir+ (1|Shrub), data=Census.Train[Census.Train$Target>0,])
summary(linmod5.lme)
drop1(linmod5.lme, test="Chisq")

linmod6.lme=lmer(LogitTarg~(DistClass+Area_Bot+Fire+TranDir)^3 - Area_Bot:Fire:TranDir - DistClass:Fire:TranDir - Fire:TranDir - DistClass:Area_Bot:TranDir+ (1|Shrub), data=Census.Train[Census.Train$Target>0,])
summary(linmod6.lme)
drop1(linmod6.lme, test="Chisq")

linmod7.lme=lmer(LogitTarg~(DistClass+Area_Bot+Fire+TranDir)^3 - Area_Bot:Fire:TranDir - DistClass:Fire:TranDir - Fire:TranDir - DistClass:Area_Bot:TranDir - DistClass:Area_Bot:Fire + (1|Shrub), data=Census.Train[Census.Train$Target>0,])
summary(linmod7.lme)
drop1(linmod7.lme, test="Chisq")

linmod7.lme=lmer(LogitTarg~(DistClass+Area_Bot+Fire+TranDir)^3 - Area_Bot:Fire:TranDir - DistClass:Fire:TranDir - Fire:TranDir - DistClass:Area_Bot:TranDir - DistClass:Area_Bot:Fire + (1|Shrub), data=Census.Train[Census.Train$Target>0,])
summary(linmod7.lme)
drop1(linmod7.lme, test="Chisq")

linmod8.lme=lmer(LogitTarg~ DistClass+Area_Bot+Fire+TranDir+ DistClass:Area_Bot + DistClass:Fire   + DistClass:TranDir + Area_Bot:Fire  +  (1|Shrub), data=Census.Train[Census.Train$Target>0,])
summary(linmod8.lme)
drop1(linmod8.lme, test="Chisq")

linmod8.lme=lmer(LogitTarg~ DistClass+Area_Bot+Fire+TranDir+ DistClass:Area_Bot + DistClass:Fire   + DistClass:TranDir + Area_Bot:Fire  +  (1|Shrub), data=Census.Train[Census.Train$Target>0,])
summary(linmod8.lme)
drop1(linmod8.lme, test="Chisq")

linmod9.lme=lmer(LogitTarg~ DistClass+Area_Bot+Fire+TranDir+ DistClass:Area_Bot + DistClass:TranDir + Area_Bot:Fire  +  (1|Shrub), data=Census.Train[Census.Train$Target>0,])
summary(linmod9.lme)
drop1(linmod9.lme, test="Chisq")


anova(linmod9.lme)


```
Model Diagnostics

Residual plot looks ok, residuals a little leptokurtic, but at least centered
```{r Model diagnostics, echo=FALSE, fig.height=3.5}
par(mfrow=c(1,3))
plot(fitted(linmod9.lme), resid(linmod9.lme), abline(c(0,0), lty=2))
qqnorm(resid(linmod9.lme))
abline(c(0,1))
hist(resid(linmod9.lme), breaks=40, freq=F)
curve(dnorm(x, mean=mean(resid(linmod9.lme)), sd=sd(resid(linmod9.lme))), add=TRUE)
#Go with this one for now???

```

```{r Presence Only fixed effects}
# Fixed effects
fixef(linmod9.lme)

```
Try on Testing Data. 
Not stochastic, uses population level data for previously unobserved levels

```{r Testing Data, echo=FALSE}


Census.Test$PresOnlyPred=NA
Census.Test[Census.Test$MHcode>1 & Census.Test$Target>0,"PresOnlyPred"]=predict(linmod9.lme,  newdata=Census.Test[Census.Test$MHcode>1& Census.Test$Target>0,], allow.new.levels=TRUE )

plot(Target/100~inv_logit(PresOnlyPred), data=Census.Test[Census.Test$MHcode>1& Census.Test$Target>0,], xlab="Predicted Percent Cover", ylab="Observed percent Cover" ) 
abline(0,1, lty=2)



```

Not Great.

```{r Save models, echo=FALSE, message=FALSE, results='hide'}
GoodPresenceModel=as.data.frame(t(fixef(hurdle12.lme)))
names(GoodPresenceModel)[1]="Int"
names(GoodPresenceModel)[which(names(GoodPresenceModel)=="PlotDist:TranDirS")]="Dist.DirS"
write.csv(as.data.frame(t(fixef(hurdle12.lme))), "ErodPresenceModel.csv", row.names=F)
GoodCoverModel=as.data.frame(t(fixef(linmod9.lme)))
names(GoodCoverModel)[1]="Int"
names(GoodCoverModel)[which(names(GoodCoverModel)=="PlotDist:Area_Bot")]="Dist.Area"
names(GoodCoverModel)[which(names(GoodCoverModel)=="Area_Bot:FireUB")]="Area.FireUB"
write.csv(GoodCoverModel, "ErodCoverModel.csv", row.names=F)
        

```

```{r plot predict results}

NewShrubs=expand.grid(PlotDist=1:250, Rain=c("A","D"), Fire=c("B","UB"), TranDir=c("N","S"), Area_Bot=mean(Census$Area_Bot), Shrub=1)

NewShrubs$PredHurd=predict(hurdle12.lme,  newdata=NewShrubs, allow.new.levels=TRUE )

NewShrubs$PredCov=predict(linmod9.lme,  newdata=NewShrubs, allow.new.levels=TRUE )

plot(NewShrubs$PlotDist, inv_logit(NewShrubs$PredHurd), col=firecols[NewShrubs$Fire])
with(NewShrubs[NewShrubs$Fire=="UB",], points(PlotDist, inv_logit(PredHurd), col=firecols[1]))



plot(Target/100~inv_logit(PresOnlyPred), data=Census.Test[Census.Test$MHcode>1& Census.Test$Target>0,], xlab="Predicted Percent Cover", ylab="Observed percent Cover" ) 
abline(0,1, lty=2)



```

